---
title: "Outline of the R session"
output: 
  html_notebook:
    toc: true
    toc_float: true
    css: style.css
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE, fig.width = 4, 
                      fig.height = 5, warning = FALSE, 
                      message = FALSE, encoding = "UTF-16")
```

## Topics
R topics we will cover in this workshop are lsit below.     

* Outlines  
* Basics of P programming  
  + About R  
  + Installation of R and Rstudio  
  + R objects  
  + Names and values  
  + R object housekeeping  
  + Recommended R programming style  
  + Getting help with R  
  + R package management  
  + R Workspace  
* Data structures  
  + Vectors
  + Lists
  + Factors
  + Matrices and arrays
  + Data frames and the like
  + More on subsetting
* Functions
  + Overview of functions 
  + Create functions
  + Invoking/calling function
  + Lexical scoping
  + Lazy evaluation
  + ... arguments
  + Exiting a function
  + Argument matching rules for function calls
  + Common functions
* Control flow
  + Choices (if-else, switch, ifelse)
  + Loops (for, while, repeat)
* I/O
  + Input
  + Output
  + Viewing data
  + Handling missing values
* Data manipulation
  + Operators
  + Tidy data with tidyr
  + Data manipulation with dplyr
  + Data structure and type conversion
* Data visualization
  + Overview of R graphics systems
  + ggplot2 and the like
* Statistical analysis
  + Univariate statistics
  + Multivariate statistics
* Running R in batch mode
  + Rscript
  + Passing arguments
  + R shell scripting
* Recommendation for further learning
  + Awesome R packages
  + Awesome R books
  + Further learning recommendations


## Install packages to be used in the workshop  
BiocManager::install() can be used to install packages from CRAN and Bioconductor sites.

```{r}
packages_list <- c("aplpack", "bootstrap", "broom", "ca", "car", "cluster", "colorspace", "ComplexHeatmap", "corrplot", "d3heatmap", "DAAG", "datasets", "dendextend", "devtools", "doBy", "downloader", "dplyr", "factoextra", "FactoMineR", "fpc", "gclus", "ggforce", "ggm", "ggthemes", "gplots", "grDevices", "grid", "gvlma", "hexbin", "Hmisc", "lattice", "leaps", "MASS", "mclust", "methods", "multcomp", "NbClust", "nFactors", "openxlsx", "pastecs", "pheatmap", "polycor", "psych", "pvclust", "Rcmdr", "RColorBrewer", "readxl", "relaimpo", "rgl", "rmarkdown", "scatterplot3d", "sm", "tibble", "tidyverse", "vcd", "vioplot", "XLConnect")

if (!require("BiocManager")) {install.packages("BiocManager")}

for (pkg in packages_list)
{
   if (!pkg %in% installed.packages())
   {
      BiocManager::install(pkg, dependencies = TRUE)
   }
}
```

To use devtools, you have to install devtools along with a working development environment.  
* Windows: Install **Rtools**.  
* Mac: Install Xcode from the Mac App Store.  
* Linux: Install a compiler and various development libraries (details vary across differnet flavors of Linux).  


## References  
1. _R in Action, Second Edition: Data analysis and graphics with R_  
By **Robert I. Kabacoff**, May 2015, ISBN 9781617291388.
Website: https://www.manning.com/books/r-in-action-second-edition

2. _Advanced R, First and Second Editions_  
By **Hadley Wickham**
Website: http://adv-r.had.co.nz/, and https://adv-r.hadley.nz/

3. Many R blogs, webpages and tutorials. For time reason, I couldn't appropriately cite them. Sorry about that.
