---
title: "Data frames and the like"
output: 
  html_notebook:
    toc: true
    toc_float: true
    css: style.css
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE, fig.width = 4, 
                      fig.height = 5, warning = FALSE, 
                      message = FALSE, encoding = "UTF-16")
```

## Data frames

A data frame is the most common way of storing data in R, making data analysis easier.  
It has properties of both a list and a matrix:  

* It is a list of vectors of the same length.  
* It likes a matrix, but different columns have different types.

### Create a data frame using _data.frame()_

```{r eval = FALSE}

data.frame(..., row.names = NULL, check.rows = FALSE,
           check.names = TRUE, fix.empty.names = TRUE,
           stringsAsFactors = default.stringsAsFactors())

# retrun TRUE
default.stringsAsFactors()
```

```{r}
# By default, strings are turned into factors
df <- data.frame(x = 1:3, y = letters[1:3], z = c(TRUE, FALSE, TRUE))
df
str(df)

# suppress data.frame()'s default behavior
df <- data.frame(x = 1:3, 
                 y = letters[1:3], 
                 z = c(TRUE, FALSE, TRUE), 
                 stringsAsFactors = FALSE)
df
str(df)
```

### Testing and coercion

```{r eval = FALSE}
df <- data.frame(x = 1:3, 
                 y = letters[1:3], 
                 z = c(TRUE, FALSE, TRUE), 
                 stringsAsFactors = FALSE)
# check type of a data frame
typeof(df)

# check its class
class(df)

# Check if an object is a data frame
is.data.frame(df)

# Coercion of other data structures into data frames

# a vector will become a one-colun data frame
a <- 1:10
as.data.frame(a)

# A list will become a data frame, with each element as an column if each list element has the same length; otherwise R throws an error
list_1 <- list(x = 1:3, 
               y = letters[1:3], 
               z = c(TRUE, FALSE, TRUE))
as.data.frame(list_1)


list_2 <- list(x = 1:3, 
               y = letters[1:3], 
               z = c(TRUE, FALSE, TRUE))
as.data.frame(list_2)

# A matrix will become a data frame with the same dimensions.
mat <- matrix(1:10, nrow = 5)
as.data.frame(mat)
```

### Attributes of a data frame
A data frame has attributes, such as names (same as col.names), row.names, and class.

Attention!!!: row.names and col.names are attributes, but rownames() and col.names() are functions.

```{r eval = FALSE}
df <- data.frame(x = 1:3, 
                 y = letters[1:3], 
                 z = c(TRUE, FALSE, TRUE), 
                 stringsAsFactors = FALSE)
# get attributes
attributes(df)

# get names, which is also column names
names(df)
colnames(df)

# get row names
rownames(df)

# get both row and column names
dimnames(df)

# get dimensions for a data frame 
nrow(df)
ncol(df)  # same as length(df)
dim(df)
```
### Subset, index and modify a data frame
subsetting, indexing and modifying a data frame can be done in both a matrix way or a list way.
Attention: subsetting will simplify the resulting data structure from a data frame to a vector if only one resulting column. 

```{r eval = FALSE}
df <- data.frame(x = 1:3, 
                 y = letters[1:3], 
                 z = c(TRUE, FALSE, TRUE), 
                 stringsAsFactors = FALSE)
# in matrix ways
df[1, ]

# see the difference with or without drop = FALSE
df[, 1, drop = FALSE]  # data frame
df[, 1]                 # vector

df[1:2, 1:2]

# in list ways
df$x
df[[1]]
df[["x"]]
df[2]

# remove a column
df[, 2] <- NULL
```

## Tibble
Tibbles are also extended and enhanced data frames, implemented by the R package **Tibble**. They tweak some older behaviors to make life a little easier. A data structure adopted by the tidyverse packages.

Key differences between tibbles and data frames are printing, subsetting, and recycling rules.  

*Column data is not coerced. A character vector is not turned into a factor. List-columns are expressly anticipated and do not require special tricks. Read more in tibble().  

Recycling only happens for a length 1 input.  

* Column names are not munged, although missing names are auto-populated. Empty and duplicated column names are strongly discouraged, but the user must indicate how to resolve. Read more in name-repair.  

* Row names are not added and are strongly discouraged, in favor of storing that info as a column. Read about in rownames.  

* df[, j] returns a tibble; it does not automatically extract the column inside. df[, j, drop = FALSE] is the default. Read more in subsetting.  

* There is no partial matching when $ is used to index by name. df$name for a nonexistent name generates a warning. Read more in subsetting.  

* Printing and inspection are a very high priority. The goal is to convey as much information as possible, in a concise way, even for large and complex tibbles. Read more in formatting.  
```{r eval = FALSE}
options(tibble.print_max = n, tibble.print_min = m, tibble.width = Inf)
```

* View full dataset with _View()_ of _glimpse()_. 

For more details, see its [vignette](https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html).

### Construct a Tibble  
```{r}
library("tibble")
# construct by columns
tb <- tibble(x = 1:3, y = letters[1:3])

# construct by rows
tb <- tribble(~x, ~y, 
        1, "a",
        2, "b",
        3, "c")
```
### Coercion between tibbles and data frames
```{r}
# conver tibbles to data frames
df <- as.data.frame(tb)  

# data frames to tibbles
tb <- as_tibble(df)

# convert named vectors to tibbles
x <- 1:3
names(x) <- letters[1:3]
x <- enframe(x)
```
### Testing tibble
```{r}
is_tibble(tb)
```
## Recommendations  

* For better performance, use tibbles, especially in case of large data.  
* In case some functions may not work with tibbles, convert tibbles to data frames using the _as.data.frame(). 

