---
title: "Summary of subsetting"
output: 
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Subsetting operators in R are powerful and fast. Mastery of subsetting allows you to succinctly express complex operatins in a way that few other languages can match.  
> --Hardly Wickham  

In this section, we will comprehensively study:  

* Six type data can be used used for subsetting.  
* Important differences in behaviors for different objects (vectors, lists, factors, matrices, and data frames, etc).  
* Three subsetting operators: $, [ ], [[ ]].  
* The use of subsetting in conjuction with assignment.  
* Missing/out of bonds indices.  


## Subsetting atomic vectors
1. **Postive integers** returns elemetns at the specified positions.
```{r}
x <- 1:10

# re-arrange order of values
x[c(3, 1)]

# duplicate values
x[c(1,1)]
```
2. **Negative integers** omit elements at the specified positions. 
```{r}
x[-c(3, 1)]

# Don't mix positive and negative integers in a single subset
# error
try({x[c(-1, 2)]})
```
3. **logical vectors** select elements where the corresponding logical value is TRUE.
```{r}
# subsetting indices with length equal to the vector to subset
x[x > 5]

# if not, recycle the subsetting indices
x[c(TRUE, FALSE, TRUE)]
```

4. **Nothing** return the orginal data.
```{r}
x[]
```
5. **zero** returns a zero-length vector.
```{r}
x[0]
```

6. **character vectors** return elements with exactly matching names if a vector is named.
```{r}
y <- 1:10
names(y) <- letters[1:10]
y[c("a", "b", "c")]
```

## Subsetting lists
Subsetting a list works in the same way as subsetting an atomic vector.  
Attention: [ ] will always return a list; [[ ]] and $ return the object stored in a list element.  

## Subsetting matrices and arrays
You can subset higher-diemensional structures in three ways:  
1. Use **multiple vectors**, with one for each dimension
```{r}
a <- matrix(1:9, nrow = 3)
colnames(a) <- LETTERS[1:3]
a

# blank subsetting return all rows or all columns
a[1:2, ]  # equivant to a[1:2, 1:3]

# each dimension uses a different type of subsetting data

a[c(TRUE, FALSE, TRUE), c("B", "A")]

# zero index for subsetting get column names only
a[0, -2]
```

2. **A single vector** can be used for subsetting matrices because matrices/arrays are implemented as vectors with special attributes. Keep in mind, values in matrices/arrays are stored in a column-major order.

```{r}
a <- matrix(1:9, nrow = 3)
a

a[c(3,6)]
```
3. **A matrix** can be used to subset a matrices.
```{r}
a <- matrix(1:36, nrow = 6)
a
b <- matrix(c(1,1,
              2,3,  # element at row 2, col 3
              4,6), ncol = 2, byrow = TRUE)
a[b]
```
**By default, [ ] will simplfy the results to the lowest possible dimensionality**.

```{r}
a <- matrix(1:36, nrow = 6)
a[1, ]
```

## Subsetting data frames
Data frames possess the characteristics of both lists and matrices. If you subset them with one single vector, they behave like lists; if you subset with two vectors, they behave like matrics.
```{r}
df <- data.frame(x = 1:3, y = letters[1:3], z = c(TRUE, FALSE, TRUE))

# matrix-like
df[df$x < 3, ]
df[c(1,3), ]
df[, c("x", "z")]

df[, "x", drop = FALSE]  # matrix subsetting simplifies resulting data structure by default

# list-like
df[c(1,3)]
df[c("x", "z")]
df["x"]                  # But list subsetting does not.
```

## S3 objects: atomic vectors, lists, arrays, data frames and data frame-like.
Use the above subsetting methods.

## S4 objects
How to create S4 objects are specified in the R package **methods**. Many Bioconductor packages implement S4 objects.  
In S4 objects, data are stored in slots.  

Using @, equivalent to \$ for S3 objects and the _slot()_ function, equivalent to [[]]. @ is more restrictive than $. It returns an error if a slot does not 

```{r}
library(methods)

# create a class Person
setClass("Person", representation(name = "character", age = "numeric"))

# create an object of class Person
peter <- new("Person", name = "Peter", age = 31)

# get Peters's age
peter@age
```

## Susetting operators for S3 objects  
* **[ ]**, can return one or more values.  
* **[[ ]]**, similar to [ ], but it can only return a single object.  
* **$**, a shorthand for [[ ]] combined with character subsetting, returns only a single object stored at the next level of a list, or a column of a data frame as a vector.  

**Attention**: differences between **$** and **[[ ]]**  

* \$ allows partial matcing of names/colnames. x$y is equivalent to x[["y", exact = FALSE]].  
* If you store the name of column in a variable, you can't use **$** with the variable to access the column.

```{r}
list_x <- list(abc = 1)
list_x$a

# Return NULL
list_x[["a"]]

# OK
list_x[["a", exact = FALSE]]

df <- data.frame(q = 1:2, b = letters[1:2])
x <- "q"
# OK
df$q

# return NULL
df$x
```

```{r}
x <- list(a = 1, b = 3, c = 4)
x[[1]]
x[["a"]]
x$a

# Error-subscript out of bounds
try({x[[c("a", "b")]]})

# a multi-value vector can index recursively for nested lists
y <- list(a = list(b = list(c = list(d = 2))))
y[[c("a", "b", "c", "d")]] 

# same as
y[["a"]][["b"]][["c"]][["d"]]  # more explicitly, I prefer this way.
```
Data frames are special lists, so We can use [[ ]] and $ to subset them. They are also sort of more generalized matrices, wo we can using [ ] to subset them.

## Simplifying versus preserving subsetting
**Simplifying subsetting** returns the simplest possible data structure that can represent the output.

**Preserving subsetting** keep the structure of the output the same as the input. .

Attention: Subsetting matrices and data frames without drop = FALSE potentially leads to simplifying subsetting, thus potential programming errors.

When you should use **simplifying** or **preserving** subsetting? This dependends on your purpose of data manipulation. The following table provides a good summary.

| |simplifying|Preserving|
|-|----------|----------|
|Vector|x[[1]]|x[1]|
|List|x[[1]]|x[1]|
|Factor|x[1:4, drop = TRUE]|x[1:4]|
|Array|x[1, ] or x[, 1]|x[1, , drop = FALSE] or x[, 1, drop = FALSE]|
|Data frame|x[, 1] or x[[1]]|x[, 1, drop = FALSE] or x[1]|

* Atomic vectors: remove names if it is named
```{r}
x <- c( a= 1, b = 2, c = 3)
x[1]
x[[1]]
```
* List: return the object inside the list, not a singl element list.
```{r}
y <- list(a = 1, b = 2)
str(y[1])
str(y[[1]])
```

* Factor: drop any unused levels
```{r}
z <- factor(c("a", "b"))
z[1]
z[1, drop = TRUE]
```
* Matrix or array: if any of the dimensions has length of 1, drop that dimension.
```{r}
a <- matrix(1:4, nrow = 2)
a[1, ]
a[1, , drop = FALSE]
```
* Data frame: if outpur is a singl column, returns a vector instead of a data frame.
```{r}
df <- data.frame(a = 1:2, b = letters[1:2])
str(df[1])
str(df[[1]])
str(df[, "a", drop = FALSE])
str(df[, "a"])
```

## Subsetting and assignment
All subsetting operators can be combined with assignment to modify selected values of the input data. WE have seen some of these operation before.

```{r}
x <- 1:5
x[1:2] <- 2:3
x

# The length of the replacement must be the same as the part to be replaced
# OR the length of the replacement must be one
x[-1] <- 4:1
x

# You can't combine integer indices with NA
# Error
try({x[c(1, NA)] <- 1:2})

# But you can combine logical indices with NA, which is coerced to FALSE as indices
x[c(TRUE, FALSE, NA)] <- 1
x

# When compare NA to anything, you get NA
df <- data.frame(a = c(1, 10, NA))
df$a[df$a < 5] <- 0
df$a

# Subsetting with nothing can be useful in conjuction with assignment, because it preserve the original object class and structure.
# compare the following two statements
mtcars[] <- lapply(mtcars, as.integer)
str(mtcars)

mtcars_1 <- lapply(mtcars, as.integer)
str(mtcars_1)

# Subsetting + assignment for list
# remove a list element

x <- list(a = 1, b = 2)
x[["b"]] <- NULL
str(x)

y <- list(a = 1)

# pay attention to how to subset a non-existing list element
y["b"] <- list(NULL)
str(y)
# What will you get if you do as follows?
y <- list(a = 1)
y$b <- NULL

str(y)

y[["b"]] <- NULL
str(y)

y[["b"]] <- list(NULL)

str(y)
```

## Exercise  
Fix each of the following common data frame subsetting errors:
```{r eval = FALSE}
str(mtcars)
mtcars[mtcars$cyl = 4, ]
mtcars[-1:4, ]
mtcars[mtcars$cyl <= 5]
mtcars[mtcars$cyl == 4 | 6, ]
```