---
title: "Running R in Batch Mode"
output: 
  html_notebook:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
So far we have run R using Rstudio IDE in an interactive way. Many times we need to run R non-interactively (aka, Batch mode) if the process is long and computing-intensive. To see how to run R in batch mode, check the R manual.
```bash
R --help
``` 
R can be run in non-interactive mode in several ways. All of them involve
working from the command line or some terminal emulator. In Mac systems,
you use the __terminal__. In Windows systems you use the __command prompt__,
or _PowerShell_, or any other shell alternative (e.g. GitBash, Cygwin).

Assuming that you have an R script file `myscript.R` that will 
be executed non-interactively, you can use one of the following options:

```bash
# Using 'R CMD BATCH' 
R CMD BATCH myscript.R
# Using 'Rscript'
Rscript myscript.R
# Using input redirection
R < myscript.R
# Running a shell script
./myscript.R
```

## Using R CMD BATCH or Rscript  
This is modified from a [R blog](https://www.r-bloggers.com/passing-arguments-to-an-r-script-from-command-lines/) and a [GitHub document](https://github.com/gastonstat/tutorial-R-noninteractive/blob/master/03-rscript.Rmd)

- `Rscript` is the alternative front-end of `R CMD BATCH`
- output goes to standard output (_stdout_) by default
- `Rscript` has default options `--slave --no-restore`

The usage of `Rscript` has the following form:
```bash
Rscript [options] [-e expr [-e expr2 ...] | file] [args]
```
- `options` are options beginning with double dash `--`. These can be any of 
the standard `R` front-end.
- `expr` (further expressions `expr2 ...`) refers to any R expression, and must be properly quoted.
- `file` is the name of the input file containing R commands
- `args` are arguments to be passed to the script in `file`

You can run more than one expression using as many `-e` flags as
expressions to be executed (make sure to properly quote them)
```bash
Rscript -e '2 + 3' -e '2 * 3' -e '2 ^ 3'
```

You can also execute a compound expression with multiple commands 
separated with semicolons `;`
```bash
Rscript -e '2 + 3; 2 * 3; 2 ^ 3'
```

Here's one example of how to render an `.Rmd` (R markdown) file from the
command line (again, be careful with the quotes):

```bash
Rscript -e 'library(rmarkdown); render("document.Rmd")'
```

### Passing arguments to R script

The common way to pass arguments to R scripts from command lines is using _Rscript_ or _R CMD BATCH_ with **trailing** arguments. For example, 

```bash
Rscript --vanilla toyScript.R geneList.txt out.txt
```

The common way to extract arguments passed to R scripts from command lines is to use the _commandArgs()_ function. See its [documentation](https://www.rdocumentation.org/packages/R.utils/versions/2.9.0/topics/commandArg). 

For example, let create a toy example R script name _toyScript.R_
```{r eval = FALSE}
#!/usr/bin/env Rscript

## the first element of args is the first argument, instead of the name of the file containing the R script itself.
args <- commandArgs(trailingOnly = TRUE)   

# test if there is at least one argument: if not, return an error
if (length(args) == 0) {
  stop("At least one argument must be supplied (input file).n", call. = FALSE)
} else if (length(args) == 1) {
  # default output file
  args[2] <- "out.txt"
}

df = read.table(args[1], header = TRUE)
num_vars = which(sapply(df, class) == "numeric")
df_out = df[, num_vars]
write.table(df_out, file = args[2], row.names = FALSE)
```

Make _toyScript.R_ executable by changing its permission as follows: 
```bash
chmod +x toyScript.R
```
Then run it as follows:  
```bash
toyScript.R input.data.frame.txt output.data.frame.txt
```


Another example:  
Save the first code snippet as "normal-vector.R".  

```bash
#!/usr/bin/env Rscript
# Generates a random vector following a normal distribution.
# Values for 'n', 'mean', 'sd' are expected (in that order).
# reading arguments ('n', 'mean', 'sd')
args <- commandArgs(trailingOnly = TRUE)
n <- as.numeric(args[1])
mean <- as.numeric(args[2])
sd <- as.numeric(args[3])
x <- rnorm(n, mean, sd)
cat(x, '\n')
```
Assuming that you change the file permissions (chmod +x normal-vector.R), you can run it as:
```bash
./normal-vector.txt 10 5 1
```

Another way is to use _--args_ option of the R command
You can also pass arguments with the `--args` options of the `R` command:

```bash
R --vanilla --args 10 5 2 < normal-vector.R
```

## R Shell Script  
In this case we use a `bash` shell that instead of using the command `Rscript` to interpret the script, it uses the `R` command. The file is `bash-example.sh`

```bash
#!/bin/bash
# creates a csv file with log-transformed
# variable 'mpg' and 'hp' from data set 'mtcars'
R --slave <<EOF
log_mtcars <- function()
{
	log_mpg <- log(mtcars$mpg)
	log_hp <- log(mtcars$hp)
	logdata <- data.frame(log_mpg, log_hp)
	write.csv(logdata, "log-mtcars.csv")
}
log_mtcars()
EOF
```
Now, make this file executable by
```bash
chmod +x bash-example.sh
```
To run the R function included in this file type
```bash
./bash-example.sh
```

Another way to execute R scripts in the command line is with 
the `R` command and the redirection operator `<`.

Consider the file `myscript1.R`, which generates two random vectors, fit a regression line, and generates two plots:
```{r regression, eval = FALSE}
# Regression analysis of two random vectors
# random data
x <- rnorm(20)
y <- x + rnorm(20)
# regression line
reg <- lm(y ~ x)
# scatter diagram with fitted regression line
png('scatterplot.png')
plot(x, y, las = 1, pch = 19, col = "#555555")
abline(reg, col = "#0000DD59", lwd = 2)
dev.off()
# residuals plot
png('residuals_plot.png')
plot(x, reg$residuals, las = 1, pch = 19, col = "#606060")
abline(h = 0)
dev.off()
```
Here's how to execute this script in the command line with a redirection 
operator:  

```bash
R --vanilla < myscript1.R
```